---
title: "Exploratory Data Analysis on Crime Data"
output: html_notebook
---

This analysis explores crime data from the Los Angeles Police Department (LAPD), covering incidents reported between January 1, 2020, and March 27, 2025. This data is transcribed from original crime reports that are typed on paper and therefore there may be some inaccuracies within the data. Some location fields with missing data are noted as (0°, 0°). Address fields are only provided to the nearest hundred block in order to maintain privacy. This data is as accurate as the data in the database. Despite these limitations, the data is considered as accurate as the information stored in the LAPD database. This report employs exploratory data analysis (EDA) using R to uncover patterns, trends, and insights.

**About this Dataset:**

Date Created: February 10, 2020

Data Last Updated: March 27, 2025

Data Provided by: Los Angeles Police Department

Dataset Owner: LAPD OpenData

Refresh rate: Bi-monthly

Category: Public Safety

Source Link: <http://data.lacity.org>

### Install and load libraries for data manipulation, visualization, and analysis.

```{r}
## Install required packages if not already installed

# Load libraries
library(tidyverse)  # For data manipulation and visualization
library(lubridate)  # For date/time handling
library(skimr)      # For quick data summaries
library(janitor)    # For cleaning column names
library(leaflet)    # For mapping geographic data
```

### Load the Data

```{r}
##Read the CSV file into R. The dataset has columns like DR_NO, Date Rptd, DATE OCC, AREA NAME, Crm Cd Desc, etc., as described in the documentation.

# Load the dataset
crime_data <- read_csv("Crime_Data_from_2020_to_Present_20250409.csv")

# Clean column names (make them consistent, remove spaces)
crime_data <- clean_names(crime_data)

head(crime_data)
```

### Initial Data Inspection

Check the structure, summary, and missing values to understand the dataset’s quality and characteristics.

```{r}
# Check structure of the dataset
str(crime_data)

# Quick summary of the dataset
summary(crime_data)
```

```{r}
# Use skimr for a detailed overview
skim(crime_data)
```

```{r}

# Check unique values in categorical columns

unique(crime_data$area_name)
unique(crime_data$crm_cd_desc)
unique(crime_data$vict_sex)
unique(crime_data$vict_descent)
```

### Data Cleaning

Handle missing values, incorrect data types, and anomalies based on the dataset’s characteristics.

```{r}
# Create a copy
crime_data_copy1 <- crime_data
```

```{r}
# Convert date columns to proper date format
crime_data_copy1$date_rptd <- mdy_hms(crime_data_copy1$date_rptd)

# Convert time_occ to a proper time format (it's in 24-hour military time as text)
crime_data_copy1$time_occ <- sprintf("%04d", as.numeric(crime_data_copy1$time_occ))  # Pad with zeros
crime_data_copy1$time_occ <- hm(paste0(substr(crime_data_copy1$time_occ, 1, 2), ":", substr(crime_data_copy1$time_occ, 3, 4)))

# Create separate columns for month, day, year and day of the week by date reported
crime_data_copy1 <- crime_data_copy1 %>%
  mutate(day=day(date_rptd),
         month=month(date_rptd, label=TRUE),
         year=year(date_rptd),
         day_of_week=wday(date_rptd, label=TRUE))

# Extract date components
crime_data_copy1 <- crime_data_copy1 %>%
  mutate(hour = hour(time_occ))
```

```{r}
# Replace (0, 0) coordinates with NA
crime_data_copy1 <- crime_data_copy1 %>%
  mutate(lat = na_if(lat, 0),
         lon = na_if(lon, 0))

# Handle missing or unknown values in categorical columns
crime_data_copy1 <- crime_data_copy1 %>%
  mutate(vict_sex = na_if(vict_sex, "X"),
         vict_sex = na_if(vict_sex, "H"),
         vict_sex = na_if(vict_sex, "-"),
         vict_descent = na_if(vict_descent, "X"),
         vict_descent = na_if(vict_descent, "-"))

# Convert vict_age to numeric, handle non-numeric or negative values
crime_data_copy1$vict_age <- as.numeric(crime_data_copy1$vict_age)
crime_data_copy1$vict_age[crime_data_copy1$vict_age <= 0] <- NA

# Check for duplicates based on DR_NO (should be unique)
duplicates <- crime_data_copy1[duplicated(crime_data_copy1$dr_no), ]
if (nrow(duplicates) > 0) {
  print("Duplicates found:")
  print(duplicates)
} else {
  print("No Duplicates found")
}
head(crime_data_copy1)
```

```{r}
# Use skimr for a detailed overview
skim(crime_data_copy1)
```

```{r}
## Identify the column names with complete_rate >= 0.73
# Calculate the skim statistics for the dataset
skim_output <- skim(crime_data_copy1)

# Get the complete_rate for each column
# The complete_rate is stored in the 'complete_rate' column of the skim_df
# The variable names are in the 'skim_variable' column
complete_rates <- skim_output$complete_rate
variable_names <- skim_output$skim_variable


# Identify the names of columns to keep (complete_rate >= 0.73)
columns_to_keep <- variable_names[complete_rates >= 0.73]
crime_data_filtered <- crime_data_copy1[, columns_to_keep]

# Remove unneccessary columns
#columns_to_remove <- crime_data_filtered
  #cdro("date_occ","mocodes","crm_cd_1")
crime_data_final_columns <- crime_data_filtered[,!names(crime_data_filtered) %in% c("date_occ","mocodes","crm_cd_1")]

crime_data_cleaned <- drop_na(crime_data_final_columns)
head(crime_data_cleaned)
```

```{r}
skim(crime_data_cleaned)
```

### Univariate Analysis

Explore individual variables to understand their distributions.

```{r}
# Numerical variables: vict_age
ggplot(crime_data_cleaned, aes(x = vict_age)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  labs(title = "Distribution of Victim Age", x = "Age", y = "Count") +
  theme_minimal()

# Crime type: crm_cd_desc
crime_data_cleaned %>%
  count(crm_cd_desc) %>%
  top_n(10, n) %>%
  ggplot(aes(x = reorder(crm_cd_desc, n), y = n)) +
  geom_bar(stat = "identity", fill = "red") +
  coord_flip() +
  labs(title = "Top 10 Crime Types", x = "Crime Description", y = "Count") +
  theme_minimal()

```

### Temporal Analysis

Analyze crime patterns over time using date_occ and time_occ.

```{r}
# Crimes by month
crime_data_cleaned %>%
  count(month) %>%
  ggplot(aes(x = month, y = n)) +
  geom_bar(stat = "identity", fill = "purple") +
  labs(title = "Crimes by Month", x = "Month", y = "Count") +
  theme_minimal()


# Crimes by day of the week
crime_data_cleaned %>%
  count(day_of_week) %>%
  ggplot(aes(x = day_of_week, y = n)) +
  geom_bar(stat = "identity", fill = "blue") +
  labs(title = "Crimes by Day of Week", x = "Day", y = "Count") +
  theme_minimal()

```

### Victim Demographics

Explore vict_age, vict_sex, and vict_descent.

```{r}
# Victim sex distribution

library(scales) # Load the scales package

crime_data_cleaned %>%
  count(vict_sex) %>%
  ggplot(aes(x = vict_sex, y = n, fill = vict_sex)) +
  geom_bar(stat = "identity") +
  # Add this layer to control the y-axis formatting
  scale_y_continuous(labels = label_number()) +
  labs(title = "Victim Sex Distribution", x = "Sex", y = "Count") +
  theme_minimal()

# Victim descent
crime_data_cleaned %>%
  filter(!is.na(vict_descent)) %>%
  count(vict_descent) %>%
  ggplot(aes(x = reorder(vict_descent, n), y = n)) +
  geom_bar(stat = "identity", fill = "darkgreen") +
  # Add this layer to control the y-axis formatting
  scale_y_continuous(labels = label_number()) +
  coord_flip() +
  labs(title = "Victim Descent Distribution", x = "Descent", y = "Count") +
  theme_minimal()
```

### Conclusion

-   Crime patterns vary by time, location, and demographics

-   Theft and assault are prevalent crime types

-   Data limitations require careful handling
